\thetitle

### 2020-11-14 writing/mutate.tex
revision of mutate

\hline
  \texttt{quadT} & Converter from S-expresion source code to PDF format (M. Butterick)\\%[1em]

\hline
  \texttt{quadU} & Converter from S-expresion source code to PDF format  (B. Greenman) \\%[1em]

@@

\hline
  \texttt{synth} & Converter of description of notes and drum beats to WAV format (V. St-Amour \& N. Toronto) \\%[1em]

@@

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow
~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
A mutation is a local syntactic change of the code of a program that may
produce a bug.
\ll{Need citations for mutation}
Usually, the outcome of mutating a program
once, dubbed a mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming errors, and mutation frameworks typically come with standard mutators carefully designed to create such mutants.

However, Typed Racket's gradual type system detects shallow errors, unlike those introduced by standard mutators.
Thus, we need different mutators to evaluate the rational programmer, ones which are fine-tuned to generate shallow errors.
The table in figure~\ref{table:mutation-ops} summarizes our mutators.
For each one, the table provides a short description and an example
mutation it performs. Each mutator is inspired by the more-than-a-decade
long experience of the authors making type-level mistakes in Typed Racket.
We find that these mistakes often take non-trivial effort to debug.

Most of our mutators in figure~\ref{table:mutation-ops} are self-explanatory.
The first twelve can apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
\texttt{Positive-Integer}. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
aim to confuse occurrence typing.

@@

\subsection{Are these mutators good enough?}

Of course, mutators that introduce these kinds of shallow errors directly into the benchmarks do not produce bugs that are interesting from the perspective of the rational programmer.
The benchmarks are fully typed, so Typed Racket's type checker locates them immediately.
For that reason, we apply the mutators to versions of the benchmarks with some components stripped of type annotations (dubbed \emph{configurations}).
Without the annotations, the type checker may not detect the errors in these mutant configurations.
Instead, Typed Racket's gradual type system may detect the errors dynamically.
Mutant configurations that result in dynamic errors are the ideal ground to exercise the rational programmer's ability to locate errors.

Specifically, we define an \emph{interesting debugging scenario} as a mutant configuration that
\begin{enumerate}
\item would be rejected by the type checker if we restored the missing type annotations,
\item produces a dynamic error under Erasure as-is, and
\item the stacktrace of the error mentions at least three distinct components.
\end{enumerate}

The second and third criteria require a little more explanation.
The second criterion requires that the error is detectable under Erasure to ensure that it is detectable by all three semantics.
This is a direct consequence of~\citet{gfd-oopsla-2019}'s theorem that a dynamic type error under Erasure implies a dynamic type error under both Transient and Natural.
It is worth noting that this criterion excludes from consideration all scenarios where Natural or Transient can detect the bug but Erasure cannot, and thus it favors Erasure.
However, our goal here is to evaluate blame across these three semantics, and thus we focus on systems where all of them exhibit blame.

The third criterion aims to capture debugging scenarios with a specific kind of dynamic error: one produced by the interactions between typed and untyped components.
In those cases, the rational programmer may need to examine several components to locate the source of the bad interaction.
In technical terms, we conservatively approximate these scenarios by analyzing the stacktrace of the error.
If the stack trace mentions at least two components, then this is evidence of interaction between components.
We require the third component because our benchmarks come with a driver component which is present in every stacktrace.

@@

With this definition in hand, we analyze how effectively our mutators produce interesting debugging scenarios from our benchmarks.
Figure~\ref{fig:mutant-breakdown} shows the number of distinct interesting debugging scenarios produced by our mutators for each benchmark.
The columns correspond to our benchmarks, and each rows show the count of interesting debugging scenarios produced by one mutator from figure~\ref{table:mutation-ops}.
The counts are cut of at 50, so any bar reaching the right side of the plot means 50 or more scenarios.

Figure~\ref{fig:mutant-breakdown} illustrates that our mutators do indeed produce a large number of interesting debugging scenarios for all of our benchmarks.
All of the mutators together produce at least 40 interesting scenarios for every benchmark, and those scenarios are produced by at least four different mutators for each benchmark.
Thus our mutators produce a large and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one benchmark.
Some mutators clearly only target one or two of them, though, and in particular the class-focused mutators are primarily effective in \texttt{take5}.
This reflects that among our benchmarks \texttt{take5} makes the most extensive use of object-oriented features, while the other features targeted by the rest of the mutators are used more broadly.

>>>>

\hline
  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]

\hline
  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]

@@

\hline
  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]

@@

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
inject bugs with mutation
analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
mutation is a local syntactic change of the code of a program that may
produce a bug. Usually, the outcome of mutating a program once, dubbed a
mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming
errors, and mutation frameworks typically come with standard mutators
carefully designed to create such mutants.

However,  unlike the deep errors introduced by standard mutators, 
gradual type systems detect shallow errors.  Thus, in order to
evaluate the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to generate shallow errors.  Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

Most of our mutators are self-explanatory.
The first twelve apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
\texttt{Positive-Integer}. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
aim to confuse occurrence typing.

@@

\subsection{Are these mutators any good?}

The astute reader must have observed that mutators that cause interesting
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking flags each impedance at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{configurations}, of the benchmarks where some components are
stripped of their type annotation.  Without the annotations, the type
checker may not detect an impedance and, instead, the type-level checks
of the gradual type system signal a run time error.  Mutant configurations
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

Specifically, we define an \emph{interesting debugging scenario} as a
mutant configuration that
\begin{enumerate} 
  \item is rejected by the type checker when we restore any missing type annotations, 
  \item produces a run time error under Erasure as-is, and 
  \item the stack trace of the error mentions at least three distinct components.  
\end{enumerate}

The first criterion simply validates that the debugging scenario concerns
a type-level mistake. However, The second and third criteria require a
little more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
theorem that a run time type error under Erasure implies a run time type
error under both Transient and Natural.  It is worth noting that this
criterion excludes from consideration all scenarios where Natural or
Transient can detect an impedance but Erasure cannot, and thus it favors
Erasure.  However, our goal here is to test the rational programmer across
these three semantics, and thus we opt for a setting where all of them
notice the impedance.

The third criterion aims to capture debugging scenarios with a specific
kind of run time error: one produced by the interaction of a buggy
component with the rest of the program.  In those cases, the rational programmer
may need to examine several components to locate the source of the faulty interaction.
In technical terms, we conservatively mark these scenarios by
analyzing the stack trace of a run time error.  If the stack trace mentions at
least two distinct components, then this is evidence of interaction between
components.  We require a third component because our benchmarks come
with a driver component which is present in the stack trace of all run
time errors.

@@

With this definition in hand, we analyze how effective our mutators
are to turn the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. 
Figure~\ref{fig:mutant-breakdown} breaks down the interesting
debugging scenarios produced  for each benchmark bu mutator 
but counts are cut off at 50. The mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a large and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark and some mutators clearly only target one or two of them, which
justifies their inclusion. For
instance the class-focused mutators are primarily effective in
which jives with the fact that \texttt{take5} is the benchmark with 
the most extensive use of object-oriented features, while the other
features targeted by the rest of the mutators are used more broadly.
